The effects of binarity on planet occurrence rates measured by transit surveys

### 2017-11-28 sandbox/note_km.tex
km note: changed mu defn. Wrote Model 2.

\title{ The effects of binarity on planet occurrence rates measured by transit 
surveys}
%
\correspondingauthor{L. Bouma}
\email{luke@astro.princeton.edu}
%
\author{L. G. Bouma}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
\author{J. N. Winn}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
\author{K. Masuda}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
%
%
\begin{abstract}
%
This derives the equations of the paper.

@@

We can detect a signal if 
\begin{equation}
	{\text{signal}\over\text{noise}}\sim
    {\delta_{\rm obs}\over(L/d^2)^{-1/2}},
	\quad \delta_{\rm obs}:\text{observed depth}
\end{equation}
is above some threshold (it would probably make more sense to include the 
duration information, but that would anyway be a trivial extension and so we 
omit it here for brevity). Thus, the maximum searchable distance scales as
\begin{equation}
	d(\delta_{\rm obs}, L_{\rm sys}) \propto \delta_{\rm obs} \cdot L_{\rm 
	sys}^{1/2}.
\end{equation}
We assume that the signal is detected if and only if a given star is searchable.

@@

Given the apparent signal $\delta_{\rm obs}$ and stellar mass $M\a$, the 
maximum searchable distance for singles and binaries are proportional to 
$\delta_{\rm obs}\cdot L(M\a)^{1/2}$ and $\delta_{\rm obs}\cdot 
[L(M\a)+L(qM\a)]^{1/2}$. Thus, 
the apparent number of searchable stars (i.e. points in the sky), which will 
be selected by an ignorant observer, can be written 
\begin{align}
	\notag
	N_{\rm s,a}(\delta_{\rm obs}, M\a)
	%&=n_{\rm s,0}\delta_{\rm obs}^3L(M\a)^{3/2} + n_{\rm b,0}\delta_{\rm 
	%obs}^3 \cdot [L(M\a)+L(qM\a)]^{3/2}\\
	&\propto n_{\rm s}\delta_{\rm obs}^3 L(M\a)^{3/2}\left[ 1+\int 
	\mathrm{d}q\,f(q) 
	{\mathrm{BF}\over{1-\mathrm{BF}}}\left(1+{L(qM\a) \over 
	L(M\a)}\right)^{3/2}\right]\\
	N_{\rm s,a}(\delta_{\rm obs}, M\a)
    &\equiv N\s^0(\delta_{\rm obs}, L(M\a)) \left[1+\mu(\mathrm{BF}, 
	M\a)\right],
\end{align}
where $N\s^0$ is the number of searchable singles (this agrees with the actual 
value), $f(q)$ is the binary mass ratio distribution, $\mathrm{BF}=n_{\rm 
b}/(n_{\rm s}+n_{\rm b})$ is the binary fraction in a volume-limited sample, 
and $n_s$ is the number density of singles in a volume-limited sample.
Proportionality constants subsumed into the definition of $N_s^0$ include 
{\it e.g.}, the telescope area and the survey duration.

@@

Since we assume $\ps\a=\ps_1$,
\begin{equation}
	\mathcal{N}_{\rm s}^1(r_a, M_a; r, M) \propto \delta(M_a-M).
\end{equation}
In this case, $\mathcal{N}_{\rm s}^1$ is non-zero only at 
$r_a=R_a\sqrt{\delta_{\rm obs}}$, 
and the observed depth is
\begin{equation}
	\delta_{\rm obs}
	= \left[{r\over R(M_a)}\right]^2\times {L(M_a) \over L_{\rm sys}(M_a, q)}.
   \label{eq:delta_obs_primaries} 
\end{equation}
The normalization of $\mathcal{N}_{\rm s}^1$ is given by the number of 
binaries that are searchable for a signal $\delta_{\rm obs}$:
\begin{equation}
	N_{\rm s}^0(\delta_{\rm obs}, 
	L(M_a))\cdot\mu(\mathrm{BF},M_a).
\end{equation}
Thus, the number of primaries with apparent parameters $(r_a,M_a)$ given the 
true parameters $(r,M)$ is
\begin{equation}
	\mathcal{N}_{\rm s}^1(r_a, M_a; r, M)
	=\int \mathrm{d}q\,f(q)\mathcal{N}_{{\rm s}, q}^1(r_a, M_a; r, M; q),
\end{equation}
where $f(q)$ is the binary mass ratio distribution and
\begin{align}
	\notag
	\mathcal{N}_{{\rm s}, q}^1(r_a, M_a; r, M; q)
	&=N_{\rm s}^0(\delta_{\rm obs}, L(M_a))\cdot\mu(\mathrm{BF},M_a)\\
	&\times\delta \left(r_a-r\sqrt{{L(M_a) \over L_{\rm sys}(M_a, 
	q)}}\right)\delta(M_a-M).
\label{eq:N_s_1}
\end{align}
Note that while the observed depth is a function of the true parameters 
(Eq.~\ref{eq:delta_obs_primaries}), since we are counting 
detected planets of a given {\it apparent} size, we will not need to worry 
about this dependence.

@@

\subsubsection{Simplifying above general result}
To simplify, drop the explicit $M_a$ and BF 
dependences, and assume $R\propto M$, so $R(qM_a)/R(M_a) = q$.
Also, move all superscript numbers to subscripts.
Then the apparent occurrence rate is
\begin{equation}
\Lambda\a(r\a) = \frac{1}{1+\mu}
\left(
    \Lambda_0(r\a) +
    \mu \left[
        \int \frac{{\rm d}q}{\mathcal{A}(q)} f(q) 
        \Lambda_1\left( \frac{r\a}{\mathcal{A}(q)} \right)
        +
        \int \frac{q {\rm d}q}{\mathcal{B}(q)} f(q)
        \Lambda_2\left(\frac{r\a}{\mathcal{B}(q)}\right) q^{2/3}
    \right]
\right).
\end{equation}
For the power law case when $L\sim M^\alpha$,
\begin{equation}
\mathcal{A}(q)
=(1+q^\alpha)^{-1/2},
\quad
\mathcal{B}(q)
=q^{-1}(1+q^{-\alpha})^{-1/2}
\end{equation}

@@

\subsubsection{Same Planets}

If $\Lambda_i(r, M)=Z_i\cdot\delta(r-r_0)$ (all the planets have the same 
radius), by using the identity $\delta(ax+b)=\delta(x+b/a)/|a|$,
\begin{align}
	\Lambda\a(r\a)
	={1 \over 1+\mu(\mathrm{BF})}
	\left[Z_0 \cdot \delta(r\a-r_0)+(Z_1+Z_2)\cdot \sqrt{2} 
	\mu(\mathrm{BF})\cdot 
	\delta\left(r\a-\frac{r_0}{\sqrt{2}}\right)\right].
    \label{eq:model1_apparent_rate}
\end{align}
This reproduces Eq.(9) of the draft (after removing the $p_{\rm det}$ terms).
In the limit of $\mathrm{BF}\to1$ ($\mu\to\infty$), the above formula yields 
$\Gamma\a=(Z^1+Z^2)\,\delta(r\a-{r_0/\sqrt{2}})$.

@@

If we assume 
\begin{equation}
	L(M) \sim M^\alpha \sim R^\alpha,
\end{equation}
we find
\begin{align}
	A=(1+q^\alpha)^{-1/2}, \quad 
	B=q^{-1}(1+q^{-\alpha})^{-1/2},
\end{align}
so
\begin{align}
	\Lambda\a(r\a, M\a)=...
	%&=(1-\mathrm{BF})\cdot\Lambda^0(r\a, M\a)\\
	%&+\mathrm{BF}\cdot 
	%\int \mathrm{d}q\,f(q)
	%(1+q^\mathcal{A})^2 \cdot \Lambda^1\left( r\a\sqrt{1+q^\mathcal{A}}, 
	%M\a\right)\\
	%&+\mathrm{BF}\cdot
	%\int {\mathrm{d}q}\,f(q)q^{8/3}
	%\sqrt{1+q^{-\mathcal{A}}}(1+q^\mathcal{A})^{3/2}
	%\cdot\Lambda^2\left(r\a q\sqrt{1+q^{-\mathcal{A}}}, qM\a\right).
\end{align}
We may further assume
\begin{equation}
	f(q)\sim q^\beta, \quad \Lambda(r) \sim r^\gamma
\end{equation}
and keep calculating...

%\acknowledgements
%It was a pleasure to share discussions with Kento Masuda, who pointed us in 
%this direction, and helped clarify that something like this was worth 
%studying.
%It was a pleasure discussing this work with F. Dai and T. Barclay.

>>>>

\title{ The effects of binarity on planet occurrence rates measured by transit 
surveys}
%
\correspondingauthor{L. Bouma}
\email{luke@astro.princeton.edu}
%
\author{L. G. Bouma}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
\author{J. N. Winn}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
\author{K. Masuda}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
%
%
\begin{abstract}
%
This note formally derives a general equation for the apparent occurrence rate 
used in the paper.

@@

We can detect a signal if 
\begin{equation}
	{\text{signal}\over\text{noise}}\sim
    {\delta_{\rm obs}\over(L/d^2)^{-1/2}},
	\quad \delta_{\rm obs}:\text{observed depth}
    \label{eq:snr_limited}
\end{equation}
is above some threshold (it would probably make more sense to include the 
duration information, but that would anyway be a trivial extension and so we 
omit it here for brevity). Thus, the maximum searchable distance scales as
\begin{equation}
	d(\delta_{\rm obs}, L_{\rm sys}) \propto \delta_{\rm obs} \cdot L_{\rm 
	sys}^{1/2}.
\end{equation}
We assume that the signal is detected if and only if a given star is searchable.

@@

Given the apparent signal $\delta_{\rm obs}$ and stellar mass $M\a$, the 
maximum searchable distance for singles and binaries are proportional to 
$\delta_{\rm obs}\cdot L(M\a)^{1/2}$ and $\delta_{\rm obs}\cdot 
[L(M\a)+L(qM\a)]^{1/2}$. Thus, 
the apparent number of searchable stars (i.e. points in the sky), which will 
be selected by an ignorant observer, is proportional to
\begin{align}
	\notag
N_{\rm s,a}(\delta_{\rm obs}, M\a)
&\propto n_{\rm s}\delta_{\rm obs}^3 L(M\a)^{3/2}\left[1+\mu(\mathrm{BF}, 
M\a)\right],
\end{align}
and we can define
\begin{align}
	N_{\rm s,a}(\delta_{\rm obs}, M\a)
    &\equiv N\s^0(\delta_{\rm obs}, L(M\a)) \left[1+\mu(\mathrm{BF}, 
	M\a)\right],
    \label{eq:N_selected}
\end{align}
where $N\s^0$ is the number of searchable singles (this agrees with the actual 
value), $\mathrm{BF}=n_{\rm b}/(n_{\rm s}+n_{\rm b})$ 
is the binary fraction in a volume-limited sample, 
and $n_s$ is the number density of singles in a volume-limited sample.
Proportionality constants subsumed into the definition of $N_s^0$ include 
{\it e.g.}, the telescope area and the survey duration.

\subsubsection{What is $\mu$?}
In Eq.~\ref{eq:N_selected}, $\mu$ is the ratio of the number of single to 
double star systems (that are searchable for an observed signal $\delta_{\rm 
obs}$).
If we make various simplifying assumptions, it turns out that this number 
depends only on the binary fraction and the 
binary mass ratio distribution in a volume-limited sample.
The proof is as follows.

In a shell of thickness ${\rm d}r$, there are $n_d(r) 4\pi r^2 {\rm d}r$ 
searchable double star systems.
The number density of searchable double star systems $n_d(r)$ is constant up 
to the maximum searchable distance for singles, $d_0$. At greater distances, 
$n_d(r)$ decreases, because only some binaries have sufficient flux to make 
the SNR cut.
Specifically,
\begin{equation}
n_d(r) = 
\begin{cases}
n_b,	\quad r\leq d_0\\
n_b \int_{q_{\rm min}(r)}^1 {\rm d}q\, f(q), \quad d_0 < r < \sqrt{2}d_0 \\
\end{cases}
\end{equation}
where $f(q)$ is the mass ratio distribution in a volume-limited sample, 
$q_{\rm min}(r)$ is the minimum mass ratio required to provide sufficient flux 
for searchability.
Assuming $L\propto M^\alpha$,
\begin{equation}
q_{\rm min}(r) =
\left(
    \left( \frac{r}{d_0}\right)^2
    - 1
\right)^{\frac{1}{\alpha}}.
\end{equation}
Using the common parametrization $f(q) \sim q^\beta$, and setting $\beta=0$,
\begin{equation}
n_d(r) = 
\begin{cases}
n_b,	\quad r\leq d_0\\
n_b (1-q_{\rm min}(r)), \quad d_0 < r < \sqrt{2}d_0. \\
\end{cases}
\end{equation}
Integrating,
\begin{align}
N_d = \int {\rm d} N_d &= \int n_d(r) 4\pi r^2 {\rm d}r \\
&=
4\pi n_b \left(
    \frac{d_0^3}{3} +
    \left[\frac{r^3}{3}\right]^{\sqrt{2 d_0}}_{d_0}
    -
    \int_{d_0}^{\sqrt{2}d_0} \left(
    \left(\frac{r}{d_0}\right)^2 -1
    \right)^{\frac{1}{\alpha}}
    r^2 \,{\rm d}r
\right) \\
&=
\frac{4 \pi n_b}{3} d_0^3 \left(
2^{3/2} -
3 \int_{1}^{\sqrt{2}} {\rm d}u\, u^2 (u^2 - 1)^{1/\alpha}
\right), \label{eq:dimensionless_integral}
\\
&\equiv
\frac{4 \pi n_b}{3} d_0^3 \left(
2^{3/2} - 3 I
\right),
\end{align}
where the penultimate line substituted $u=r/d_0$.
For the common case $\alpha=3.5$, the dimensionless integral $I$ evaluates to 
$\approx 0.484174$.

Thus
\begin{equation}
\mu \equiv \frac{N_d}{N_s^0}
= \frac{n_b}{n_s} \left(2^{3/2} - 3 I\right)
= \frac{{\rm BF}}{1 - {\rm BF}} \left(2^{3/2} - 3 I\right).
\label{eq:mu_final}
\end{equation}
For the twin binary case, $I=0$, and for the power law volume-limited 
distribution, $I$ takes the form given in Eq.~\ref{eq:dimensionless_integral}.

@@

Since we assume $\ps\a=\ps_1$,
\begin{equation}
	\mathcal{N}_{\rm s}^1(r_a, M_a; r, M) \propto \delta(M_a-M).
\end{equation}
In this case, $\mathcal{N}_{\rm s}^1$ is non-zero only at 
$r_a=R_a\sqrt{\delta_{\rm obs}}$, 
and the observed depth is
\begin{equation}
	\delta_{\rm obs}
	= \left[{r\over R(M_a)}\right]^2\times {L(M_a) \over L_{\rm sys}(M_a, q)}.
   \label{eq:delta_obs_primaries} 
\end{equation}
The normalization of $\mathcal{N}_{\rm s}^1$ is given by the number of 
binaries that are searchable for a signal $\delta_{\rm obs}$:
\begin{equation}
	N_{\rm s}^0(\delta_{\rm obs}, 
	L(M_a))\cdot\mu(\mathrm{BF},M_a).
\end{equation}
Thus, the number of primaries with apparent parameters $(r_a,M_a)$ given the 
true parameters $(r,M)$ is
\begin{equation}
	\mathcal{N}_{\rm s}^1(r_a, M_a; r, M)
	=\int \mathrm{d}q\,f(q)\mathcal{N}_{{\rm s}, q}^1(r_a, M_a; r, M; q),
\end{equation}
where $f(q)$ is the binary mass ratio distribution in the SNR-limited sample 
and
\begin{align}
	\notag
	\mathcal{N}_{{\rm s}, q}^1(r_a, M_a; r, M; q)
	&=N_{\rm s}^0(\delta_{\rm obs}, L(M_a))\cdot\mu(\mathrm{BF},M_a)\\
	&\times\delta \left(r_a-r\sqrt{{L(M_a) \over L_{\rm sys}(M_a, 
	q)}}\right)\delta(M_a-M).
\label{eq:N_s_1}
\end{align}
Note that while the observed depth is a function of the true parameters 
(Eq.~\ref{eq:delta_obs_primaries}), since we are counting 
detected planets of a given {\it apparent} size, we will not need to worry 
about this dependence.
As discussed in the main text, we have $f(q) \sim q^\beta (1+q^\alpha)^{3/2}$.

@@

\subsubsection{Simplifying above general result}
To simplify, drop the explicit $M_a$ and BF 
dependences, and assume $R\propto M$, so $R(qM_a)/R(M_a) = q$.
Also, move all superscript numbers to subscripts.
Then the apparent occurrence rate is
\begin{equation}
\Lambda\a(r\a) = \frac{1}{1+\mu}
\left(
    \Lambda_0(r\a) +
    \mu \left[
        \int \frac{{\rm d}q}{\mathcal{A}(q)} f(q) 
        \Lambda_1\left( \frac{r\a}{\mathcal{A}(q)} \right)
        +
        \int \frac{q {\rm d}q}{\mathcal{B}(q)} f(q)
        \Lambda_2\left(\frac{r\a}{\mathcal{B}(q)}\right) q^{2/3}
    \right]
\right).
\label{eq:simple_apparent_occ_rate}
\end{equation}
For the power law case when $L\sim M^\alpha$,
\begin{equation}
\mathcal{A}(q)
=(1+q^\alpha)^{-1/2},
\quad
\mathcal{B}(q)
=q^{-1}(1+q^{-\alpha})^{-1/2}
\label{eq:powerlaw_A_B}
\end{equation}

@@

\subsubsection{Same Planets (Model \# 1)}

If $\Lambda_i(r, M)=Z_i\cdot\delta(r-r_0)$ (all the planets have the same 
radius), by using the identity $\delta(ax+b)=\delta(x+b/a)/|a|$,
\begin{align}
	\Lambda\a(r\a)
	={1 \over 1+\mu(\mathrm{BF})}
	\left[Z_0 \cdot \delta(r\a-r_0)+(Z_1+Z_2)\cdot 
	\mu(\mathrm{BF})\cdot 
	\delta\left(r\a-\frac{r_0}{\sqrt{2}}\right)\right].
    \label{eq:model1_apparent_rate}
\end{align}
This reproduces Eq.(9) of the draft (after removing the $p_{\rm det}$ terms).
In the limit of $\mathrm{BF}\to1$ ($\mu\to\infty$), the above formula yields 
$\Lambda\a=(Z^1+Z^2)\,\delta(r\a-{r_0/\sqrt{2}})$.

@@

If we assume 
\begin{equation}
	L(M) \sim M^\alpha \sim R^\alpha,
\end{equation}
we find 
\begin{equation}
\mathcal{A}(q)
=(1+q^\alpha)^{-1/2},
\quad
\mathcal{B}(q)
=q^{-1}(1+q^{-\alpha})^{-1/2}.
\end{equation}

This gives a value of $\mu({\rm BF})$ stated in Eq.~\ref{eq:mu_final}.
Now recall Eq.~\ref{eq:simple_apparent_occ_rate} for the apparent occurrence 
rate. We restate it here:
\begin{equation}
\Lambda\a(r\a) = \frac{1}{1+\mu}
\left(
\Lambda_0(r\a) +
\mu \left[
\int \frac{{\rm d}q}{\mathcal{A}(q)} f(q) 
\Lambda_1\left( \frac{r\a}{\mathcal{A}(q)} \right)
+
\int \frac{q {\rm d}q}{\mathcal{B}(q)} f(q)
\Lambda_2\left(\frac{r\a}{\mathcal{B}(q)}\right) q^{2/3}
\right]
\right).
\end{equation}

\subsubsection{Same Planets (Model \# 2)}
For our ``Model \#2'', we again want $\Lambda_i(r) = Z_i \cdot \delta(r-r_0)$.
To achieve this, write
\begin{equation}
f(q) = \frac{q^\beta (1+q^\alpha)^{3/2}}{\mathcal{N}_q},
\end{equation}
for $\mathcal{N}_q$ the normalization.
For $i=1$, call the integral on the left the $[\ldots]$ ``$I_1$''.
Then
\begin{align}
I_1(r_a) &\equiv \int \frac{{\rm d}q}{\mathcal{A}(q)} f(q) 
    \Lambda_1\left( \frac{r\a}{\mathcal{A}(q)} \right)
    \\
&= \frac{1}{\mathcal{N}_q} \int {\rm d}q\, q^\beta (1+q^\alpha)^2
    \Lambda_1(r_a (1+q^\alpha)^{1/2})
    \\
&= \frac{Z_1}{\mathcal{N}_q} \int {\rm d}q\, q^\beta (1+q^\alpha)^2
    \delta(r_a (1+q^\alpha)^{1/2} - r_0) \\
&= \frac{Z_1}{\mathcal{N}_q}
    \left( \left(\frac{r_0}{r_a}\right)^2 - 1  \right)^{\beta/\alpha}
    \left( \frac{r_0}{r_a}\right)^4.
    \label{eq:model_2_i_1}
\end{align}
The penultimate line used the fact that $r_a(1+q^\alpha)^{1/2}=r_0$ has only 
one root:
\begin{equation}
q(r_a) = \left(\left(\frac{r_0}{r_a}\right)^2 - 1\right)^{1/\alpha}.
\end{equation}
Eq.~\ref{eq:model_2_i_1} should be used as a sanity check on any 
Monte Carlo numerics.

Attempting the same thing for $i=2$,
\begin{align}
I_2(r_a) &\equiv \int \frac{q {\rm d}q}{\mathcal{B}(q)} f(q)
\Lambda_2\left(\frac{r\a}{\mathcal{B}(q)}\right) q^{2/3}
\\
&= \frac{Z_2}{\mathcal{N}_q} \int {\rm d}q\,
    q^{\beta+\frac{8}{3}} 
    (1+q^\alpha)^{3/2}
    (1+q^{-\alpha})^{1/2}
    \delta(r_a q (1+q^{-\alpha})^{1/2} - r_0).
\end{align}
Unfortunately, analytic roots of $r_a q (1+q^{-\alpha})^{1/2} = r_0$ do not 
exist.
Schematically, since for any function $g(x)$,
\begin{equation}
\delta(g(x)) = \sum_{x_i \in \{{\rm roots}\}} \frac{\delta(x-x_i)}{|g'(x_i)|},
\end{equation}
if the roots are found numerically then
\begin{equation}
I_2(r_a) = \frac{Z_2}{\mathcal{N}_q}
    \sum_{q_i \in \{{\rm roots}\}}
q_i^{\beta+\frac{8}{3}} 
(1+q_i^\alpha)^{3/2}
(1+q_i^{-\alpha})^{1/2}.
\end{equation}
But this doesn't really give much insight.

\subsubsection{Power law planets (Model \# 3)}

Assume now that
\begin{equation}
    \Lambda(r) \sim r^\delta
\end{equation}
and keep calculating...
